---
title: "houston_blackout"
format: pdf
editor: visual
execute: 
  warning: false
  message: false
---

---
title: Spaitial Analysis of the 2021 Power Outage in Houston, Texas
author: William Mullins
date: 11-10-2025
---

```{r, warning = FALSE, message = FALSE}
reqired_packages <- c("sf", "terra", "tidyverse", "tmap", "here", "stars", "knitr", "units")

for(pkg in reqired_packages){
  if (!require(pkg, character.only = TRUE)){
     stop(paste0("Package '", pkg, "' is not installed"))
  } 
}

# Load required packages
suppressPackageStartupMessages({
  library(sf)
  library(tmap)
  library(tidyverse)
  library(here)
  library(knitr)
  library(patchwork)
  library(paletteer)
  library(units)
})
```

### Data Import

```{r, warning = FALSE, message = FALSE}
# Feb 7 Data
nl_1_07 <- rast(here("data", "VNP46A1", "VNP46A1.A2021038.h08v05.001.2021039064328.tif"))
nl_2_07 <- rast(here("data", "VNP46A1", "VNP46A1.A2021038.h08v06.001.2021039064329.tif"))

# Feb 16 Data
nl_1_16 <- rast(here("data", "VNP46A1", "VNP46A1.A2021047.h08v05.001.2021048091106.tif"))
nl_2_16 <- rast(here("data", "VNP46A1", "VNP46A1.A2021047.h08v06.001.2021048091105.tif"))

# Read In cencus geometry data
census_data_geometry <- st_read(dsn = "data/ACS_2019_5YR_TRACT_48_TEXAS.gdb", 
                       layer =  "ACS_2019_5YR_TRACT_48_TEXAS") %>% 
  st_transform(crs = 3083) # Change to desired crs

# Combine nightlight data by date
# Data from 2021-02-07
feb_7_comb <- mosaic(nl_1_07, nl_2_07) 
# Data from 2021-02-16
feb_16_comb <- mosaic(nl_1_16, nl_2_16) 
```

### Comparison of the Beofre and of the Outage

```{r, warning = FALSE, message = FALSE}
plot_one_data <- c(feb_7_comb, feb_16_comb)
names(plot_one_data) <- c("Radiance \n(nW cm²/sr)", "Other")
tm_shape(plot_one_data) +
 tm_graticules(crs = 4326, labels.show = TRUE) +
 tm_raster(col.scale = tm_scale_intervals(values = "-viridis",
                                          style = "quantile", )) +
  tm_title("Nighttime Light in Houston Texas Before and During the Outage") +
  tm_facets(ncol = 2, 
            free.scales = FALSE, 
            drop.units = TRUE) +
  tm_layout(panel.labels = c("February 7, 2021", "February 16, 2021"),
            legend.position = tm_pos_out("left", "center", pos.v = "center"))
```

### Identifying Sites With Blackouts
```{r, warning = FALSE, message = FALSE}
# Find the difference in light between the two days
light_change <- feb_16_comb - feb_7_comb

# Set non-significant areas to NA
light_change[light_change <= 200] <- NA

light_change_vector <- terra::as.points(light_change, na.rm = TRUE)

blackouts <- light_change_vector %>% 
  st_as_sf() %>%# Converts to an sf
  st_make_valid() %>% # Remove NA rows
  st_transform(3083) # Change to desired CRS


# Create bounding box for Houston
custom_bbox <- c(xmin = -96.5, xmax = -94.5, ymin = 29, ymax = 30.5)
# Convert convert bbox to a sf object
bbox_sfc <- st_bbox(custom_bbox) %>% 
  st_as_sfc() %>%
  st_set_crs(4326) %>%
  st_transform(3083) 

# Filter blackout data to be within the bounding box
blackouts_intersection <- st_intersects(blackouts, bbox_sfc, sparse = FALSE)
blackouts_cropped <- blackouts[blackouts_intersection,]
```


### Exclude Sites Within 200 Meters of a Road
```{r, warning = FALSE, message = FALSE}
# Road Data
roads <-read_sf(here::here("data", "gis_osm_roads_free_1.gpkg"), 
                query = "SELECT * FROM gis_osm_roads_free_1 WHERE fclass='motorway'"
)

if (st_crs(roads) != st_crs(blackouts_cropped)) {
  roads <- st_transform(roads, crs = st_crs(blackouts_cropped)) 
}

# Change the road data to have a 200m buffer
road_buffer <- st_buffer(st_union(st_geometry(roads)), dist = set_units(200, "m"))

# Create a logical vector of the intersections
close_to_roads <- st_intersects(
  st_geometry(blackouts_cropped), 
  road_buffer,
  sparse = FALSE
) %>%
  rowSums() %>% # Get the number of intersections
  as.logical() # Store as a True/False value

# Return data frame where there was not intersection
blackouts_roads <- blackouts_cropped[!close_to_roads,]

# Check that the CRS match before plotting
if(st_crs(census_data_geometry) != st_crs(blackouts_roads)) {
  message("Warning! CRS do onot match")
}
# Plot the areas that experienced blackouts
tm_shape(census_data_geometry, bbox = bbox_sfc) +
  tm_graticules(crs = 3083, labels.show = TRUE) +
  tm_borders(fill = "lightgrey",
             lwd = 0.5) +
  tm_shape(blackouts_roads) +
  tm_polygons(col = "#008080") + 
  tm_compass(position = c("right", "bottom"), size = 0.5) +
  tm_title(text = "Areas Affected by February 2021 Blackout") 
```

### Locating Where Homes Experienced Blackouts

```{r, warning = FALSE, message = FALSE}
# Houses
houses <- read_sf(here::here("data", "gis_osm_buildings_a_free_1.gpkg"), query = "
    SELECT *
    FROM gis_osm_buildings_a_free_1
    WHERE (type IS NULL AND name IS NULL)
       OR type IN ('residential', 'apartments', 'house', 'static_caravan', 'detached')
  ")

# Check if CRS match. If not then make them match
if (st_crs(houses) != st_crs(blackouts_roads)) {
  houses <- st_transform(houses, crs = st_crs(blackouts_roads)) 
}

# Data frame of houses that had blackouts.
blackouts_houses <- st_join(houses, blackouts_roads, join = st_intersects, left = FALSE)
```


```{r, warning = FALSE, message = FALSE}
# Plot the houses which experienced a blackout
tm_shape(census_data_geometry, bbox = bbox_sfc) +
  tm_graticules(crs = 3083, labels.show = TRUE) +
  tm_borders(fill = "lightgrey",
             lwd = 0.5) +
  tm_shape(blackouts_houses) +
  tm_dots(fill = "#008080",
          size = 0.5) + 
  tm_compass(position = c("right", "bottom"), size = 0.5) +
  tm_title(text = "Houses Which Experienced Power Outages February 2021") 
```

### Estimate of the Number of Houses With Outages

```{r, warning = FALSE, message = FALSE}
estimate <- blackouts_houses %>%  
 st_drop_geometry() %>%
 filter(fclass == 'building') %>% 
 summarize(home_count = n())
```

An estimated 211 homes where affected by the blackout

### Socioeconomic Variables

```{r, warning = FALSE, message = FALSE}
# Census income data
suppressMessages(
census_data_income <- st_read(dsn = "data/ACS_2019_5YR_TRACT_48_TEXAS.gdb", 
                       layer =  "X19_INCOME")
 )

# Keep columns containing the GEOID and the median income
median_income_data <- census_data_income %>%
  dplyr::select("GEOID", "B19013e1") 

# Left join geometry and income data by their respective GEOID columns
median_income_data <- dplyr::left_join(
  census_data_geometry, 
  median_income_data, 
  
  by = c("GEOID_Data" = "GEOID")
)

# Get tracts where there was an outage
tracts_with_blackouts <- st_join(median_income_data,
                                 blackouts_roads,
                                 join = st_intersects,
                                 left = FALSE)
```

### Site Classification
```{r, warning = FALSE, message = FALSE}
# Create new column and assign "no" to it as a default
median_income_data$blackout <- "No"

# Go through each row and check if there was a  blackout
for (i in 1:nrow(median_income_data)) {
  current_row <- median_income_data$TRACTCE[i]
  
  # Check if this ID exists in new_data
  if (current_row %in% tracts_with_blackouts$TRACTCE) {
    median_income_data$blackout[i] <- "Yes"
  }
  # Otherwise it remains "no" (already initialized)
}
```

### Census Tracts Which Experienced Outages

```{r, warning = FALSE, message = FALSE}
custom_palette <- c("#f5deb3", "#008080")

tm_shape(median_income_data, bbox = bbox_sfc) + 
  tm_polygons("blackout",
              fill.scale = tm_scale(values = custom_palette, 
                                    labels = c("No", "Yes")),
              fill.legend = tm_legend(title = "Blackout \n Tract", 
                                      title.size = 1,
                                      text.size = .7),
              lwd = 0.1,
              ) +
  tm_compass(position = c("right", "bottom"), size = 0.5) +
  tm_title('Cencus Tracts With Blackouts in Houston')
```

### Box PLot of Median Income vs. Blackout Occcurance
```{r, warning = FALSE, message = FALSE}
ggplot(data = median_income_data, aes(x = blackout, y = B19013e1)) +
  scale_fill_manual(values = custom_palette,  # Colorblind-safe colors
    labels = c("No Blackout", "Blackout")) +
  geom_boxplot(fill = custom_palette) +
  labs(
    title = "Median Household Income by Blackout Status",
    x = "Power Outage",
    y = "Median Household Income (USD)"
  ) + 
  theme_minimal(base_size = 14)
```

### Reflection 
Our results suggest that there seems to be a fairly small amount of correlation between the median income of people and whether or not they experienced a power outage at that time. This does not exclude the possibility of income playing a role in power outages, however, as the methods used to approximate this are limited. Rather than examining the median income of the houses directly, the median income of entire tracts is being used. This means that if just one household has a blackout during this time, its entire census tract will be reported as having experienced outages. In the future, it would be better to examine the proportion of tracts that experienced outages compared to the median income in that area.
